Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-08-18 latex/main.tex
Updates from Shing

The summary (abstract) should consist of a single paragraph of 150 words or fewer.

@@

ancestral recombination graphs, recombination, SARS-CoV-2

@@

The genomic revolution has brought a paradigm shift in how we study the evolution and spread of pathogens, 
allowing real-time tracking of pathogens over time and space. 
During the COVID-19 pandemic, thousands of genome sequences of SARS-CoV-2 samples were generated daily, 
quickly overwhelming typical phylogenetic approaches. 
Although real-time phylogenetic methods now exist to cope with this flood of genomic data 
(e.g., UShER\cite{Turakhia2021,DeMaio2023}), we lack methods to build phylogenies 
that can integrate recombination in the face of ongoing sampling of infected cases.

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies 
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}. 
UShER has generated phylogenies updated daily \cite{McBroome2021} 
and proved to be a powerful resource that successfully enabled the world 
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}). 
These phylogenies are used to identify and monitor 
potentially epidemiologically important “Pango” lineages \cite{Rambaut2020,McBroome2024}, 
which might have acquired features to become Variants of Concern (VOC), 
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology, 
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2 
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}). 
UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}. 
Given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2022}, 
the tree-like representation of UShER is no longer tenable.

@@

Sc2ts reconstructs an ARG by incrementally adding batches of sequences 
as they become available (Figure 1A; for details, see STAR Methods). 
First, sc2ts infers likely “copying paths” connecting each sample in the batch to node(s) in the current ARG 
(allowing for recombination) using the Li and Stephens (LS) model \cite{Li2003}, 
which is a Hidden Markov Model (HMM) widely used in human genomics 
to approximate the effects of mutation and recombination on genetic inheritance. 
As recombination is relatively rare in viruses, 
we would expect most samples to copy from a single parent in the ARG. 
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path 
between the parents resulting in different genomic segments inherited from the different parents. 
Next, for clusters within the batch that are inferred to copy from the same ARG node, 
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster. 
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, 
and applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

Sc2ts uses collection dates to build a time-resolved ARG. 
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions), 
likely due to errors in their collection dates 
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1, 
which is estimated to have emerged in November, 2021; REF). 
To preclude adding such “time travelling” samples, we filtered samples by “HMM cost”. 
We define the “HMM cost” of a sample to be the total penalty cost 
for switches and mismatches (mutations) in its copying path. 
We parameterise the LS model such that each mismatch has a penalty cost of one 
and a switch has a penalty cost of \textit{k} (here, set to 4), 
which determines how much mutation is favored over recombination when accounting for a sample (Figure 1B). 
If the HMM cost of a sample exceeds a threshold (here, set to 7), 
sc2ts does not immediately add the sample to the ARG.

This approach resulted in preliminary ARGs that had noticeably fewer time travellers, 
but we found that it also excluded the major “saltation” events that characterize many VOCs. 
These saltation lineages are characterised by a high number of accumulated mutations, 
and are hypothesised to emerge from chronic infections 
during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}. 
Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron). 
To distinguish time travellers with erroneous dates from true saltation lineages, 
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added. 
If on subsequent days these lineages form a sufficiently large group of samples, 
sc2ts adds the group back to the ARG. 
This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX). 
The emergence of the Delta, BA.1 and BA.2 lineages required some manual intervention, 
in which specific "seed" samples were identified for unconditional inclusion 
(for details of how seed samples were chosen, see STAR Methods). 
First arising in geographic regions with low sampling density, 
the initial origins of these VOC are challenging to infer (REF).

\noindent\includegraphics[width=0.85\linewidth]{methodology.png}

\subsection*{Figure 1: Schematic overview of sc2ts.} 
(A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates. 
Sample 0 is used to initialise the ARG as the root. 
Next, Sample 1 is attached, mapping mutations onto the new branch to explain it. 
When Sample 4 is detected as a recombinant, 
a node representing the recombination event is inserted with two edges connecting to its parents, 
with Sample 4 itself attached to this node as its child. 
The resulting ARG contains two local trees, one on each side of the inferred breakpoint. 
(B) Copying paths for Sample 4 inferred under the LS model 
while allowing recombination (\textit{k} = 4) or disallowing it (\textit{k} = 1000).

@@

The ARG represents the full evolutionary history of SARS-CoV-2 
and is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian data \cite{Hunt2024}. 
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024} 
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods). 
Tanglegrams reveal that the ARG captures known phylogenetic relationships both within sublineages 
and at the level of major lineages (including VOCs), 
with good agreement between the phylogenetic bipartitions in the ARG and the UShER phylogeny (Figure S2). 
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian data 
by pruning both to their intersection of 2,475,418 samples and 27,431 sites. 
Remarkably, given the differences in inference methods and data processing pipelines, 
they are almost identically parsimonious, 
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts). 
We also compared the mutational spectra of major VOCs in the ARG 
against the mutational spectra computed from the UShER phylogeny (STAR Methods), 
finding a close match to previous reports \cite{Bloom2023} (Figure S3).

Input genome alignments are masked in sc2ts 
so that deletions and ambiguous bases are treated as missing data by the HMM, 
and once the samples are incorporated into the ARG, 
these bases are imputed to carry the values inherited by the node in question. 
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however, 
and to incorporate them into the ARG we identified a set of [XXX] sites 
most frequently involved in deletions [cite source?] 
and post-hoc mapped alignment data for these sites to the ARG using parsimony. 
Although sites are treated independently in this approach, 
the approximation successfully identified the most frequent deletions, 
such as the recurrent 21991-21993 (see Document S1 for more details). 
The node representing the origin of Alpha, with 290,407 descendants, 
is correctly associated with key lineage-defining deletions: 
11288-11296 (ORF1A del3675-3677), 
21765-21770 (Spike del69-70), 
21991-21993 (Spike del144), and 
28271 (non-coding). 
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021}, 
also emerges twice in the origin of Omicron - 
in the node leading to the BA.1 variant (342,099 descendants), 
and the node leading to the BA.4 variant (189,746 descendants). 
To speed up inference and to avoid issues caused by sites enriched for errors [citation?], 
sc2ts masks out a set of 100 sites identified as the most highly homoplasic 
by running on a subset of the data (STAR Methods). 
Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER [citation?]. 
In addition to remapping the data for sites identified as being involved in deletions, 
we added mutations for XX of these highly recurrent sites 
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony. 
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods). 
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment 
(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata). 
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG, 
increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).

@@

As new data became available during the pandemic, 
volunteers continually detected recombinants which may develop concerning features, 
particularly increased transmissibility, disease severity, 
and ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2022}). 
The process of gathering evidence for and characterising such recombinants is manual, 
involving visual inspection of phylogenies for unusual branches 
caused by forcibly placing recombinant sequences onto them, 
typically followed by visual analysis of genome sequence alignments 
to identify parental lineages and breakpoint locations. 
This process has resulted in the designation of recombinant Pango lineages 
(whose aliases begin with an “X”), to which we refer as “Pango Xs”. 
Below, we summarise our analysis of the Pango Xs in the ARG 
(for extended analyses, see Documents S1 and S2).

We find strong evidence for the recombination events leading to 17 Pango Xs in the ARG (Table 1), 
which passed the detection-level sc2ts quality control filters (Figure S1). 
These events are supported by high numbers of averted mutations (Table 1) 
and high numbers of loci matching their suggested parents (see their copying patterns in Figure SX). 
Among these Pango Xs are XA, which is the first designated Pango X showing signs of community transmission \cite{Jackson2021}, 
and XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2022}. 
The recombinant origin of XA is well explained in the ARG by two parents and one mutation (Figure 2A; Table 1), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango Xs above originate from shared recombination events in the ARG (Table 1). 
An example of a set of “nested” closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP, 
which descend from a single recombination event rather than five independent events in the ARG (Figure 2B). 
These Pango Xs were designated around the same time. 
These results highlight that the ARG can help clarify the relationships between recombinants, 
allowing for more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1), 
on the basis of the HMM and available data. 
For all these Pango Xs, the number of recombination-informative sites 
fell below our threshold of detection (i.e. k = 4 in the HMM). 
The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG 
by at most three mutations (Table 1) strongly supports their origins as non-recombinants. 
For example, the closely related XN and XAU can be explained 
by two mutations and one mutation in the ARG, respectively (Figure 2C). 
%%% TODO: Say something about XAJ and XAN.
However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants 
because the HMM treated characteristic deletions as missing data.

\noindent\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}

\subsection*{Figure 2: Illustrative local subgraphs in the ARG.}
Subgraphs around the XA-associated recombination node (depicted as an encircled R) (A); 
the recombination node associated with a group of “nested” Pango Xs (B); 
and the MRCA of the putative non-recombinants XN and XAU (C). 
For more detailed visualisations of these subgraphs, see Document S2.

@@

We compared sc2ts against three state-of-the-art (non-phylogenetic) methods 
that detect inter-lineage recombinants by scanning sequences for mutations associated with different Pango lineages: 
(1) RecombinHunt \cite{Alfonsi2024}, (2) CovRecomb \cite{Li2024}, and (3) rebar (https://github.com/phac-nml/rebar).

To evaluate the concordance of the methods in identifying the parental Pango lineages 
and locating the breakpoints of the recombinants, we focus on the Pango Xs, using the data 
from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison. 
For RecombinHunt, we took the detection results for the Pango Xs presented in the original study; 
for CovRecomb, we summarised the results from the supplementary data of the original study (STAR Methods); 
and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods). 
For each method, we determined whether the inferred parental lineages 
and breakpoint intervals are concordant with those identified by the community (STAR Methods). 
We find that the methods performed well overall, 
with no significant difference in concordance among them (Table 1; p-value = XX, chi-square test), 
indicating that sc2ts can characterise recombinants as well as the state-of-the-art recombination detection methods.

We also compared the sensitivity of sc2ts and rebar. 
Of all the “robust” recombination events detected by sc2ts, 
167 (43\%) were also detected as recombinants by rebar (STAR Methods). 
We hypothesise that this disagreement arose from differences in sensitivity between these methods. 
Consistent with this, we find that rebar needs a higher level of divergence 
between parents to correctly classify a sequence as recombinant, 
as shown by a shorter time to their MRCA (Figure 3A) 
and lower distance between their Pango identity (Figure 3B).

@@

Sc2ts provides a method to facilitate the discovery of recombination events, 
by comparing how well a sample fits the ARG under a model allowing recombination 
versus a model disallowing recombination, 
as measured by the number of resulting mutations (Figure 1B). 
The more mutations are “averted” by allowing recombination, 
the stronger the evidence for recombination. 
We calculated the number of averted mutations for all the “robust” recombination events (STAR Methods), 
which include all the Pango X-associated recombination events (Table 1).

Our analysis of the number of averted mutations reveals a spectrum of plausibility 
across all the “robust” recombination events (Figure 3C). 
Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1). 
The “copying pattern” of XA visually confirms that a high number of recombination-informative sites 
support the suggested parents (Figure 3D). 
XW is an example of a marginally well-supported recombinant. 
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent. 
The copying patterns of all the recombination events are shown in Document S3.

@@

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\noindent\includegraphics[width=0.85\linewidth]{averted_mutations.png}

\subsection{Figure 3: Evidence for recombination events in the ARG.} 
The 386 “robust” recombination events detected by sc2ts are colored 
according to their classification by rebar (recombinant: purple, non-recombinant: gray). 
(A) Distribution of time to the MRCA of the sc2ts-identified parents. 
(B) Distribution of the Pango lineage distance between the sc2ts-identified parents. 
(C) Distribution of the number of mutations averted by recombination (bottom x-axis), 
defined as the difference in the number of mutations 
between a copying path allowing recombination (right y-axis) 
and an alternative copying path disallowing recombination (top x-axis). 
(D) Copying patterns of five select recombinants.

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome, 
particularly at the left boundary of the Spike gene (Figure 4A), 
consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}. 
While this might result from a higher recombination rate in this region, 
it may also be due to the higher levels of polymorphism in the Spike gene, 
allowing recombination endpoints to be estimated more precisely. 
Reassuringly, sites having unusually high mutation counts are absent in this region (Figure 4B), 
which would confound recombination detection.

Furthermore, we find an intrinsic negative relationship 
between breakpoint interval size and the amount of divergence between recombinant parents, 
which is measured by time to their MRCA (Figure 4D). 
When the two parental lineages have diverged for longer periods of time (the rightmost points), 
more information becomes available on either side of a breakpoint to pinpoint its location, 
resulting in shorter intervals. 
For instance, the breakpoint intervals tend to be the smallest for events 
involving recombination between the deeply divergent Omicron BA.1 and BA.2 sublineages 
(the red points in Figure 4D).

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\noindent\includegraphics[width=0.85\linewidth]{recombination_landscape.png}

\subsection*{Figure 4: Recombination landscape.} 
(A) Distribution of the breakpoint intervals of the 386 “robust” recombination events along the genome 
(annotated with gene labels on top). Each line represents an interval colored by its size. 
(B) Number of overlapping breakpoint intervals (orange; left y-axis) 
and number of mutations along the genome (blue; right y-axis). 
The coordinates of the ten sites with the highest mutation counts are labelled. 
(C) Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations. 
(D) Relationship between breakpoint interval size and time 
between the recombination event and the parental MRCA. 
The black curve shows the theoretical expected relationship between these quantities, 
estimated by minimizing the sum of squared residuals given a single parameter (the mutation rate) (STAR Methods).

>>>>

%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
%%% This is the SMBE abstract. We can rewrite or work off of it, but it needs to be shorter.
Widely used phylogenetic methods to study SARS-CoV-2 evolution assume that genetic recombination is negligible and 
viral relationships can be represented by a single tree. 
However, important SARS-CoV-2 recombinant lineages have emerged. 
A more suitable approach is to use ancestral recombination graphs (ARG), 
which jointly model recombination, mutation, and genetic inheritance in a cohesive framework. 
Recent computational advances have made ARG inference possible for datasets of millions of genome sequences. 
Building on these new advances, we have developed sc2ts, a scalable method to simultaneously detect recombinants and 
reconstruct ARGs from heterochronous SARS-CoV-2 genome sequences. 
Using sc2ts, we inferred an ARG of 2.48 million genomes sampled from January, 2020 to February, 2023. 
We find that this pandemic-scale ARG unifies known features of SARS-CoV-2 evolution — 
the backbone phylogeny, mutational spectra, and major recognized recombinants. 
This ARG enables us to explore the evolutionary origins of known and novel recombinants. 
Furthermore, the ARG is compactly stored in the tree sequence format using the tskit library and 
can be analyzed efficiently and interactively using standard Python tools on a typical laptop. 
This combination of capabilities makes sc2ts a powerful tool 
for studying the recombination and mutation processes driving SARS-CoV-2 evolution.

@@

ancestral recombination graphs, recombination, tree sequence, SARS-CoV-2

@@

The genomic revolution has brought about a paradigm shift 
in the way we study the evolution and spread of pathogens, 
allowing real-time tracking of pathogens over time and space. 
During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily, 
quickly overwhelming typical phylogenetic approaches. 
Although real-time phylogenetic methods now exist to cope with this flood of genomic data 
(e.g., UShER\cite{Turakhia2021,DeMaio2023}), 
we lack methods to build phylogenies that can integrate recombination 
in the face of ongoing sampling of infected cases.

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies 
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}. 
UShER has generated phylogenies updated daily \cite{McBroome2021} and 
has proven to be a powerful resource that successfully enabled the world 
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}). 
These phylogenies are used to identify and monitor 
potentially epidemiologically important “Pango” lineages \cite{Rambaut2020,McBroome2024}, 
which might have acquired features to become Variants of Concern (VOC), 
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology, 
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2 
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}). 
UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}. 
Given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2022}, 
the tree-like representation of UShER is no longer tenable.

@@

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available 
(\ref{fig:methodology}; for details, see STAR Methods). 
First, sc2ts infers likely “copying paths” connecting each sample in the batch to node(s) in the current ARG 
(allowing recombination) using the Li and Stephens (LS) model \cite{Li2003}, 
which is a Hidden Markov Model (HMM) widely used in human genomics 
to approximate the effects of mutation and recombination on genetic inheritance. 
As recombination is relatively rare in viruses, we would expect most samples to copy from a single parent in the ARG. 
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path 
between the parents resulting in different genomic segments inherited from the different parents. 
Next, for clusters within the batch that are inferred to copy from the same ARG node, 
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster. 
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and 
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

Sc2ts uses collection dates to build a time-resolved ARG. 
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions), 
likely due to errors in their collection dates 
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1, 
which is estimated to have emerged in November, 2021; REF). 
To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”. 
We define the “HMM cost” of a sample to be the total penalty cost for switches and mismatches (mutations) in its copying path. 
We parameterize the LS model such that each mismatch has a penalty cost of one 
and a switch has a penalty cost of \textit{k} (here, set to 4), 
which determines how much mutation is favored over recombination when accounting for a sample (\ref{fig:methodology}). 
If the HMM cost of a sample exceeds a threshold (here, set to 7), sc2ts does not immediately add the sample to the ARG.

This approach resulted in preliminary ARGs that had noticeably fewer time travelers, 
but we found that it also excluded the major “saltation” events that characterize many VOCs. 
These saltation lineages are characterized by a high number of accumulated mutations, 
and are hypothesized to emerge from chronic infections 
during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}. 
Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron). 
To distinguish time travelers with erroneous dates from true saltation lineages, 
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added. 
If on subsequent days these lineages form a sufficiently large group of samples, sc2ts adds the group back to the ARG. 
This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX). 
The emergence of the Delta, BA.1 and BA.2 lineages required some manual intervention, 
in which specific "seed" samples were identified for unconditional inclusion 
(for details of how seed samples were chosen, see STAR Methods). 
First arising in geographic regions with low sampling density, 
the initial origins of these VOC are challenging to infer (REF).

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{methodology.png}
    \caption{Schematic overview of sc2ts. (A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates. Sample 0 is used to initialize the ARG as the root. Next, Sample 1 is attached, mapping mutations onto the new branch to explain it. When Sample 4 is detected as a recombinant, a node representing the recombination event is inserted with two edges connecting to its parents, with Sample 4 itself attached to this node as its child. The resulting ARG contains two local trees, one on each side of the inferred breakpoint. (B) Copying paths for Sample 4 inferred under the LS model while allowing recombination (\textit{k} = 4) or disallowing it (\textit{k} = 1000).}
    \label{fig:methodology}
\end{figure}

@@

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}. 
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024} 
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods). 
Tanglegrams reveal that the ARG captures known phylogenetic relationships 
both within sublineages and at the level of major lineages (including VOCs), 
with good agreement between the phylogenetic bipartitions in the ARG and the UShER phylogeny (Figure S2). 
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset 
by pruning both to their intersection of 2,475,418 samples and 27,431 sites. 
Remarkably, given the differences in inference methods and data processing pipelines, 
they are almost identically parsimonious, 
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts). 
We also compared the mutational spectra of major VOCs in the ARG 
against the mutational spectra computed from the UShER phylogeny (STAR Methods), 
finding a close match to previous reports \cite{Bloom2023} (Figure S3).

Input genome alignments are masked in sc2ts 
so that deletions and ambiguous bases are treated as missing data by the HMM, 
and once the samples are incorporated into the ARG, 
these bases are imputed to carry the values inherited by the node in question. 
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however, 
and to incorporate them into the ARG we identified a set of [XXX] sites 
most frequently involved in deletions [cite source?] 
and post-hoc mapped alignment data for these sites to the ARG using parsimony. 
Although sites are treated independently in this approach, 
the approximation successfully identified the most frequent deletions, 
such as the recurrent 21991-21993 (for more details, see Document S1). 
The node representing the origin of Alpha, with 290,407 descendants, 
is correctly associated with key lineage-defining deletions: 
11288-11296 (ORF1A del3675-3677), 
21765-21770 (Spike del69-70), 
21991-21993 (Spike del144), and 
28271 (non-coding). 
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021}, 
also emerges twice in the origin of Omicron - 
in the node leading to the BA.1 variant (342,099 descendants), 
and the node leading to the BA.4 variant (189,746 descendants). 
To speed up inference and to avoid issues caused by sites enriched for errors [citation?], 
sc2ts masks out a set of 100 sites identified as the most highly homoplasic 
by running on a subset of the data (STAR Methods). 
Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER \cite{Hunt2024}. 
In addition to remapping the data for sites identified as being involved in deletions, 
we added mutations for XX of these highly recurrent sites 
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony. 
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods). 
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment 
(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata). 
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG, 
increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).

@@

As new data became available during the pandemic, 
volunteers continually detected recombinants which might have concerning features, 
particularly increased transmissibility, disease severity, and 
ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2022}). 
The process of gathering evidence for and characterizing such recombinants is manual, 
involving visual inspection of phylogenies for unusual branches 
caused by forcibly placing recombinant sequences onto them, 
typically followed by visual analysis of genome sequence alignments 
to identify parental lineages and breakpoint locations. 
This process has resulted in the designation of recombinant Pango lineages 
(whose aliases begin with an “X”), to which we refer as “Pango Xs”. 
Below, we summarize our analysis of the Pango Xs in the ARG 
(for extended analyses, see Document S1 and Document S2).

We find strong evidence for the recombination events leading to 17 Pango Xs in the ARG (Table 1), 
which passed the detection-level sc2ts quality control filters (Figure S1). 
These events are supported by 
high numbers of averted mutations (Table 1) and 
high numbers of loci matching their suggested parents (see their copying patterns in Figure SX). 
Among these Pango Xs are 
XA, which is the first designated Pango X that showed signs of community transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2022}. 
The recombinant origin of XA is well explained in the ARG by two parents and one mutation (\ref{fig:subgraphs}; Table 1), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango Xs above originate from shared recombination events in the ARG (Table 1). 
An example of a set of “nested” closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP, 
which descend from a single recombination event rather than five independent events in the ARG (\ref{fig:subgraphs}). 
These Pango Xs were designated around the same time. 
These results highlight that the ARG can help clarify the relationships between recombinants, 
allowing more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1), 
on the basis of the HMM and available data. 
For all these Pango Xs, the number of recombination-informative sites 
fell below our threshold of detection (i.e. k = 4 in the HMM). 
The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG 
by at most three mutations (Table 1) strongly supports their origins as non-recombinants. 
For example, the closely related XN and XAU can be explained 
by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}). 
%%% TODO: Say something about XAJ and XAN.
However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants 
because the HMM treated characteristic deletions as missing data.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
    \caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated recombination node (depicted as an encircled R) (A); the recombination node associated with a group of “nested” Pango Xs (B); and the MRCA of the putative non-recombinants XN and XAU (C). For more detailed visualizations of these subgraphs, see Document S2.}
    \label{fig:subgraphs}
\end{figure}

@@

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art (non-phylogenetic) methods that detect inter-lineage recombinants 
by scanning sequences for mutations associated with different Pango lineages: 
(1) RecombinHunt \cite{Alfonsi2024}, 
(2) CovRecomb \cite{Li2024}, and 
(3) rebar (https://github.com/phac-nml/rebar).

To evaluate the concordance of the methods in identifying the parental Pango lineages 
and locating the breakpoints of the recombinants, we focus on the Pango Xs, using the data 
from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison. 
For RecombinHunt, we took the detection results for the Pango Xs presented in the original study; 
for CovRecomb, we summarized the results from the supplementary data of the original study (STAR Methods); 
and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods). 
For each method, we determined whether the inferred parental lineages and breakpoint intervals are 
concordant with those identified by the community (STAR Methods). 
We find that the methods performed well overall, 
with no significant difference in concordance among them (Table 2; p-value = XX, chi-square test), 
indicating that sc2ts can characterize recombinants as well as the state-of-the-art recombination detection methods.

We also compared the sensitivity of sc2ts and rebar. 
Of all the “robust” recombination events detected by sc2ts, 
167 (43\%) were also detected as recombinants by rebar (STAR Methods). 
We hypothesize that this disagreement arose from differences in sensitivity between these methods. 
Consistent with this, we find that rebar needs a higher level of divergence between parents 
to correctly classify a sequence as recombinant, as shown by 
a shorter time to their MRCA (\ref{fig:averted_mutations}) and 
lower distance between their Pango identity (\ref{fig:averted_mutations}).

@@

Sc2ts provides a method to facilitate the discovery of recombination events, 
by comparing how well a sample fits the ARG under a model allowing recombination 
versus a model disallowing recombination, as measured by the number of resulting mutations (\ref{fig:methodology}). 
The more mutations are “averted” by allowing recombination, the stronger the evidence for recombination. 
We calculated the number of averted mutations for all the “robust” recombination events (STAR Methods), 
which include all the Pango X-associated recombination events (Table 1).

Our analysis of the number of averted mutations reveals a spectrum of plausibility 
across all the “robust” recombination events (\ref{fig:averted_mutations}). 
Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1). 
The “copying pattern” of XA visually confirms 
that a high number of recombination-informative sites support the suggested parents (\ref{fig:averted_mutations}). 
XW is an example of a marginally well-supported recombinant. 
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent. 
The copying patterns of all the recombination events are shown in Document S3.

@@

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{averted_mutations.png}
    \caption{Evidence for recombination events in the ARG. The 386 “robust” recombination events detected by sc2ts are colored according to their classification by rebar (recombinant: purple, non-recombinant: gray). (A) Distribution of time to the MRCA of the sc2ts-identified parents. (B) Distribution of the Pango lineage distance between the sc2ts-identified parents. (C) Distribution of the number of mutations averted by recombination (bottom x-axis), defined as the difference in the number of mutations between a copying path allowing recombination (right y-axis) and an alternative copying path disallowing recombination (top x-axis). (D) Copying patterns of five select recombinants.}
    \label{fig:averted_mutations}
\end{figure}

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome, 
particularly at the left boundary of the Spike gene (\ref{fig:landscape}), 
consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}. 
While this might result from a higher recombination rate in this region, 
it may also be due to the higher levels of polymorphism in the Spike gene, 
allowing recombination endpoints to be estimated more precisely. 
Reassuringly, sites having unusually high mutation counts are absent in this region (\ref{fig:landscape}), 
which would confound recombination detection.

Furthermore, we find a negative relationship between 
breakpoint interval size and the amount of divergence between recombinant parents, 
which is measured by time to their MRCA (\ref{fig:landscape}). 
When the two parental lineages have diverged for longer periods of time (the rightmost points), 
more information becomes available on either side of a breakpoint to pinpoint its location, resulting in shorter intervals. 
For instance, the breakpoint intervals tend to be the smallest for events involving recombination 
between the deeply divergent Omicron BA.1 and BA.2 sublineages (the red points in \ref{fig:landscape}).

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{recombination_landscape.png}
    \caption{Recombination landscape. (A) Distribution of the breakpoint intervals of the 386 “robust” recombination events along the genome 
(annotated with gene labels on top). Each line represents an interval colored by its size. (B) Number of overlapping breakpoint intervals (orange; left y-axis) and number of mutations along the genome (blue; right y-axis). The coordinates of the ten sites with the highest mutation counts are labeled. (C) Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations. (D) Relationship between breakpoint interval size and time between the recombination event and the parental MRCA. The black curve shows the theoretical expected relationship between these quantities, estimated by minimizing the sum of squared residuals given a single parameter (the mutation rate) (STAR Methods).}
    \label{fig:landscape}
\end{figure}
